---
title: "Analysis of DIA results: Phosphoproteomics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Phosphoproteomics workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
---

```{r, echo = FALSE}
library(knitr)
```

*This script was developed for the analysis of phosphoproteomics data obtained from DIA mass spectrometry analysed with Spectronaut.* 
*The script reads in the data (PTM and proteome BGS factory reports), performs basic preprocessing, and provides an overview of the results.*
*Statistical analysis is performed with Limma.*

# Information on the study

```{r}
library(myproteomicsR)

# --- Edit these paths and parameters before running ---

# Spectronaut PTM Site BGS Factory Report (phosphoproteome)
PATH_TO_INPUT_PTM <- "path/to/ptm_site_bgs_report.tsv"
# Spectronaut Protein Group MLP Report (proteome)
PATH_TO_INPUT_Proteome <- "path/to/protein_group_mlp_report.tsv"
# Folder where the Excel output will be saved. Use "" for current directory.
PATH_TO_OUTPUT <- ""

# Minimum number of measured (non-NA) values required in at least one
# condition for a feature to be retained.
min_measured_val <- 2
# Quantile of per-run measured values used for missing value imputation
# (e.g. 0.01 = 1st percentile).
quantile_replacement_val <- 0.01
```

The path to input files are:
- `r PATH_TO_INPUT_Proteome` for the proteome.
- `r PATH_TO_INPUT_PTM` for the phosphoproteome.

Results will be saved in `r PATH_TO_OUTPUT` (adjust the path above if you want
to save the results in a specific folder).

# Load input files

```{r}
ltab <- open_phos_prot_files(
  path_to_input_protein = PATH_TO_INPUT_Proteome,
  path_to_input_ptm = PATH_TO_INPUT_PTM
)
```

There are `r nrow(ltab$proteome)` rows in the protein groups table.
There are `r nrow(ltab$ptm)` rows in the phosphorylation sites table.

# Overview of the input data

## Boxplots of quantitative values

```{r}
lapply(ltab, plot_boxplot)
```

## PCA plots

```{r}
lp <- plot_PCA(ltab$proteome, title = "Proteome", feature_name = "protein_group_accessions")
print(lp[[1]])

lp <- plot_PCA(ltab$ptm, title = "Phosphorylated sites", feature_name = "site_ID")
print(lp[[1]])
```

# Pre-processing

## Filter contaminants

```{r}
table(ltab$proteome$contaminant, ltab$proteome$filename)
table(ltab$ptm$contaminant, ltab$ptm$filename)

ltab_filtered <- lapply(ltab, function(x) {
  x %>% filter(contaminant == FALSE)
})
```

## Filter based on missing values

```{r}
ltab_filtered <- lapply(seq_along(ltab_filtered), function(x) {
  if (names(ltab_filtered)[x] == "proteome") {
    filter_value_count(input_table = ltab_filtered[[x]], 
                       feature_name = "protein_group_accessions", 
                       title = "Proteome: distribution of the number of measured values per feature and condition",
                       min_measured_val = min_measured_val)
  } else if (names(ltab_filtered)[x] == "ptm") {
    filter_value_count(ltab_filtered[[x]], 
                       feature_name = "site_ID", 
                       title = "Phosphoproteome: distribution of the number of measured values per feature and condition", 
                       min_measured_val = min_measured_val)
  }
})
names(ltab_filtered) <- c("proteome", "ptm")

g <- plot_measured_values(ltab_filtered$proteome$value_count, title = "Proteome: distribution of the number of measured values per feature and condition")
print(g)
g <- plot_measured_values(ltab_filtered$ptm$value_count, title = "Phosphoproteome: distribution of the number of measured values per feature and condition")
print(g)
```

## Check the CVs distributions

```{r}
prot.cv <- feature_cvs(input_table = ltab_filtered$proteome$filtered_table, 
                       feature_name = "protein_group_accessions")
ptm.cv  <- feature_cvs(ltab_filtered$ptm$filtered_table, 
                       feature_name = "site_ID")

g <- plot_cvs(prot.cv, title = "Proteome: distribution of the CVs per feature and condition")
print(g)
g <- plot_cvs(ptm.cv, title = "Phosphoproteome: distribution of the CVs per feature and condition")
print(g)
```

# Missing value replacement

```{r}
# Replace missing values with the per-run 1st percentile of measured values.
# Replacement is applied only when all values are missing in one condition.
ltab_replaced <- wrap_replace_missing_values(ltab_filtered, quantile_replacement_val)
```

## PCA plots

```{r}
lp <- plot_PCA(ltab_replaced$proteome, title = "Proteome before stats", feature_name = "protein_group_accessions")
print(lp[[1]])
print(lp[[2]])

lp <- plot_PCA(ltab_replaced$ptm, title = "Phosphorylated sites before stats", feature_name = "site_ID")
print(lp[[1]])
print(lp[[2]])
```

# Statistical analysis

We use Limma:
Ritchie, ME, Phipson, B, Wu, D, Hu, Y, Law, CW, Shi, W, and Smyth, GK (2015).
limma powers differential expression analyses for RNA-sequencing and microarray studies.
Nucleic Acids Research 43(7), e47.

## Make contrast matrix

```{r}
ldesign_prot <- create_limma_design(ltab_replaced$proteome)
ldesign_phos <- create_limma_design(ltab_replaced$ptm)

cat("Design and contrasts for the proteome:\n")
kable(ldesign_prot)
cat("Design and contrasts for the phosphoproteome:\n")
kable(ldesign_phos)
```

```{r}
cat("Limma for the proteome:\n")
res_prot <- run_limma(ltab_replaced$proteome, ldesign_prot$design, ldesign_prot$contrasts, feature_name = "protein_group_accessions")
cat("Limma for the phosphoproteome:\n")
res_phos <- run_limma(ltab_replaced$ptm, ldesign_phos$design, ldesign_phos$contrasts, feature_name = "site_ID")
```

# Plot Limma results

```{r}
plot_volcano(res_prot, title = "Proteome")
plot_volcano(res_phos, title = "Phosphoproteome")
```

# Save Limma results

```{r}
write_excel_output(
  unfiltered_table_proteome         = ltab$proteome,
  filtered_table_proteome           = ltab_filtered$proteome$filtered_table,
  stat_results_proteome             = res_prot,
  mv_counts_proteome                = ltab_filtered$proteome$value_count,
  proteome_after_replacement        = ltab_replaced$proteome,
  unfiltered_table_phosphoproteome  = ltab$ptm,
  filtered_table_phosphoproteome    = ltab_filtered$ptm$filtered_table,
  stat_results_phosphoproteome      = res_phos,
  mv_counts_phosphoproteome         = ltab_filtered$ptm$value_count,
  phosphoproteome_after_replacement = ltab_replaced$ptm,
  file_name = "DIA_analysis_results",
  path      = PATH_TO_OUTPUT)
```

---

```{r}
sessionInfo()
```
