---
title: "Analysis of DIA results: Phosphoproteomics"
output: html_document
date: "`r Sys.Date()`"
---

```{r, echo = F}
# Install myproteomicsR
# pak::pak("mlocardpaulet/myproteomicsR")

# load package
library(myproteomicsR)

library(knitr)
```

*This script was developed for the analysis of phosphoproteomics data obtained from DIA mass spectrometry analysed with Spectronaut.* 
*The script reads in the data (PTM and proteome BGS factory reports), performs basic preprocessing, and provides an overview of the results.*
*Statistical analysis is performed with Limma.*

# Information on the study

All the parameters for the analysis are in the file `Analysis_Parameters.R`. Please make sure to adjust the parameters in that file before running the analysis.

```{r}
source("Analysis_Parameters.R")
```

# Load input files

```{r}

ltab <- open_phos_prot_files(
  path_to_input_protein = PATH_TO_INPUT_Proteome,
  path_to_input_ptm = PATH_TO_INPUT_PTM
)

```

There are `r nrow(ltab$proteome)` rows in the protein groups table.
There are `r nrow(ltab$ptm)` rows in the phosphorylation sites table.

# Overview of the input data

## Boxplots of quantitative values

```{r}

lapply(ltab, plot_boxplot)

```

## PCA plots

```{r}
lp <- plot_PCA(ltab$prot, title = "Proteome", feature_name = "protein_group_accessions")
print(lp[[1]])

lp <- plot_PCA(ltab$ptm, title = "Phosphorylated sites", feature_name = "site_ID")
print(lp[[1]])
```

# Pre-processing

## Filter contaminants

```{r}

table(ltab$prot$contaminant, ltab$prot$filename)
table(ltab$ptm$contaminant, ltab$ptm$filename)

ltab_filtered <- lapply(ltab, function(x) {
  x[x$contaminant == FALSE,]
})

```

## Filter based on missing values

```{r}

ltab_filtered <- lapply(seq_along(ltab_filtered), function(x) {
  if (names(ltab_filtered)[x] == "proteome") {
    filter_value_count(input_table = ltab_filtered[[x]], 
                       feature_name = "protein_group_accessions", 
                       title = "Proteome: distribution of the number of measured values per feature and condition",
                       min_measured_val = min_measured_val)
  } else if (names(ltab_filtered)[x] == "ptm") {
    filter_value_count(ltab_filtered[[x]], 
                       feature_name = "site_ID", 
                       title = "Phosphoproteome: distribution of the number of measured values per feature and condition", 
                       min_measured_val = min_measured_val)
  }
})
names(ltab_filtered) <- c("proteome", "ptm")

g <-  plot_measured_values(ltab_filtered$proteome$value_count, title = "Proteome: distribution of the number of measured values per feature and condition")
print(g)
g <-  plot_measured_values(ltab_filtered$ptm$value_count, title = "Phosphoproteome: distribution of the number of measured values per feature and condition")
print(g)

```

## Check the CVs distributions

```{r}

prot.cv <- feature_cvs(input_table = ltab_filtered$proteome$filtered_table, 
                       feature_name = "protein_group_accessions")
ptm.cv <- feature_cvs(ltab_filtered$ptm$filtered_table, 
                      feature_name = "site_ID")

g <-  plot_cvs(prot.cv, title = "Proteome: distribution of the CVs per feature and condition")
print(g)
g <-  plot_cvs(ptm.cv, title = "Phosphoproteome: distribution of the CVs per feature and condition")
print(g)

```
# Missing value replacement

```{r}

# Replace missing values with 1% quantile of the measured values in the corresponding run
# Replace only when all are missing entirely in one condition

ltab_replaced <- wrap_replace_missing_values(ltab_filtered)

```

## PCA plots

```{r}
lp <- plot_PCA(ltab_replaced$proteome, title = "Proteome before stats", feature_name = "protein_group_accessions")
print(lp[[1]])

lp <- plot_PCA(ltab_replaced$ptm, title = "Phosphorylated sites before stats", feature_name = "site_ID")
print(lp[[1]])
```

# Statistical analysis

We use Limma:
Ritchie, ME, Phipson, B, Wu, D, Hu, Y, Law, CW, Shi, W, and Smyth, GK (2015).
limma powers differential expression analyses for RNA-sequencing and microarray studies.
Nucleic Acids Research 43(7), e47.

## Make contrast matrix

```{r}
ldesign <- lapply(ltab_replaced, create_limma_design)

cat("Design and contrasts for the proteome:\n")
kable(ldesign$proteome)
cat("Design and contrasts for the phosphoproteome:\n")
kable(ldesign$ptm)
```

```{r}
cat("Limma for the proteome:\n")
res_prot <- run_limma(ltab_replaced$proteome, ldesign$proteome$design, ldesign$proteome$contrasts, feature_name = "protein_group_accessions")
cat("Limma for the phosphoproteome:\n")
res_phos <- run_limma(input_data = ltab_replaced$ptm, 
                      design = ldesign$ptm$design, 
                      contrasts = ldesign$ptm$contrasts, 
                      feature_name = "site_ID")
```

# Plot Limma results

```{r}
plot_volcano(res_prot, title = "Proteome")
plot_volcano(res_phos, title = "Phosphoproteome")
```

# Save Limma results

```{r}
# Write the first data set in a new workbook
write_excel_output(
  unfiltered_table_proteome = ltab$proteome,
  filtered_table_proteome = ltab_filtered$proteome$filtered_table,
  stat_results_proteome = res_prot,
  mv_counts_proteome = ltab_filtered$proteome$value_count,
  unfiltered_table_phosphoproteome = ltab$ptm,
  filtered_table_phosphoproteome = ltab_filtered$ptm$filtered_table,
  stat_results_phosphoproteome = res_phos,
  mv_counts_phosphoproteome = ltab_filtered$ptm$value_count,
  phosphoproteome_after_replacement = ltab_replaced$ptm,
  proteome_after_replacement = ltab_replaced$proteome,
  file_name = "DIA_analysis_results",
  path = PATH_TO_OUTPUT)
```


----------------

```{r}
sessionInfo()
```