---
title: "Analysis of DIA results analysed with DIA-NN: Proteomics"
output: html_document
date: "`r Sys.Date()`"
---

```{r, echo = F}
# Install myproteomicsR
# pak::pak("mlocardpaulet/myproteomicsR")
# a specific branch:
# remotes::install_github("mlocardpaulet/myproteomicsR@PG-DIANN")

# load package
library(myproteomicsR)

library(knitr)
```

*This script was developed for the analysis of proteomics data obtained from DIA mass spectrometry analysed with DIA-NN* 
*The script reads in the data (pg_matrix.tsv), performs basic preprocessing, and provides an overview of the results (QC plots).*

# Information on the study

All the parameters for the analysis are in the file `Analysis_Parameters.R`. Please make sure to adjust the parameters in that file before running the analysis.

```{r}
source("Analysis_Parameters_proteome_diann.R")
```

# Load input files

```{r}

tab <- load_DIANN_protein_input(
  PATH_TO_INPUT_Proteome
)

```

There are `r nrow(tab)` rows in the protein groups table.
It contains `r length(unique(tab$protein_group_accessions))` protein groups.

# Overview of the input data

## Boxplots of quantitative values

```{r}

plot_boxplot(tab)

```

## PCA plot

```{r}

p <- tab %>% 
  mutate(condition = filename) %>% 
  plot_PCA(., title = "Proteome raw", feature_name = "protein_group_accessions")
print(p)

```

# Pre-processing

## Filter contaminants

```{r}

table(tab$contaminant, tab$filename)

tab_filtered <- tab[tab$contaminant == FALSE,]

```

## Filter based on missing values

```{r}

tab_filtered$condition <- rep("cond", nrow(tab_filtered))
tab_filtered <- filter_value_count(input_table = tab_filtered,
                                   feature_name = "protein_group_accessions",
                                   title = "Proteome: distribution of the number of measured values per feature and condition",
                                   min_measured_val = min_measured_val)

g <-  plot_measured_values(tab_filtered$value_count, title = "Proteome: distribution of the number of measured values per feature and condition")
print(g)

```

## Check the CVs distributions

```{r}

prot.cv <- feature_cvs(input_table = tab_filtered$filtered_table, 
                       feature_name = "protein_group_accessions")

g <-  plot_cvs(prot.cv, title = "Proteome: distribution of the CVs per feature and condition")
print(g)

```


----------------

```{r}
sessionInfo()
```